---
title: "Reproducible Research Project 1"
author: "Makiko Takahashi"
date: "Sunday, August 10, 2014"
output: html_document
---

## Assignment 1 for Reproducible Research Course on Coursera

The data set to be analyzed was obtained through the forked Github  repository.

The date of obtaining: Aug. 10, 2014

```{r}
date_downloaded <-Sys.Date()

```

This is the working directory for the assignment.

```{r}
setwd("~/GitHub/RepData_PeerAssessment1")

```

**Loading and preprocessing the data**

Load the data and take a look:

```{r}
activity <- read.table("Data/activity.csv", header =T, sep =",")

head(activity)
tail(activity)

str(activity)
```

Chnge the datatype of date column from a factor to date.

```{r}
activity$date <- as.Date(activity$date)

```

Get the summary.

```{r}
summary(activity)

```


**What is mean total number of steps taken per day?**

For now, remove all the rows with NA values.

```{r}
no.NA.activity <- activity[complete.cases(activity),]

```

Aggregate the step per day.

```{r}
steps.per.day <- aggregate(no.NA.activity$steps, by = list(no.NA.activity$date), FUN="sum")
```

Draw a histogram.

```{r}
step.hist <- hist(steps.per.day$x, breaks=10, main="The Frequency of the Step per Day", xlab="# steps")
step.hist
```

The mean and median steps per day.

```{r}
mean.steps <- mean(steps.per.day$x)
median.steps <- median(steps.per.day$x)

```

The mean is 'r mean.steps' and the median is 'r median.steps'.

**What is the average daily activity pattern?**

```{r}
## 
mean.steps.time <- aggregate(no.NA.activity$steps, by =list(no.NA.activity$interval), FUN="mean")
plot(mean.steps.time, type="l", main="Daily Activity Pattern", ylab="# steps", xlab="Interval")
```


**Imputing missing values**

The number of the rows with NA values can be calculated by subtracting the number of the row without NA values from the total number of rows in the data frame.

The number of complete cases have been calculated using complete.case function earlier and stored in no.NA.activity variable.

Therefore, the number of the rows with NA values can be calculated as such:

```{r}
num.NA <- nrow(activity) - nrow(no.NA.activity)
num.NA
```


```{r}
days1 <- unique(activity$date)
days2 <- unique(no.NA.activity$date)
length(days1)
length(days2)
```

There are some days that are missing all the steps values.

**Are there differences in activity patterns between weekdays and weekends?**


